---
jupyter:
  jupytext_format_version: '1.0'
  jupytext_formats: ipynb,Rmd:rmarkdown,R
  kernelspec:
    display_name: R
    language: R
    name: ir
  language_info:
    codemirror_mode: r
    file_extension: .r
    mimetype: text/x-r-source
    name: R
    pygments_lexer: r
    version: 3.5.1
  toc:
    base_numbering: 1
    nav_menu: {}
    number_sections: true
    sideBar: true
    skip_h1_title: true
    title_cell: Table of Contents
    title_sidebar: Contents
    toc_cell: false
    toc_position: {}
    toc_section_display: true
    toc_window_display: true
---

R regression modeling exercise solutions
============

Exercise solutions<span class="tag" data-tag-name="prototype"></span>
=====================================================================

Exercise 0 prototype
--------------------

Use the *states.rds* data set.

```{r, eval = FALSE}
  states <- readRDS("dataSets/states.rds")
```

Fit a model predicting energy consumed per capita 
(energy) from the percentage of residents living in 
metropolitan areas (metro). Be sure to

1.  Examine/plot the data before fitting the model

```{r, eval = FALSE}
  states.en.met <- subset(states, select = c("metro", "energy"))
  summary(states.en.met)
  plot(states.en.met)
  cor(states.en.met, use="pairwise")
```

1.  Print and interpret the model `summary`

```{r, eval = FALSE}
  mod.en.met <- lm(energy ~ metro, data = states)
  summary(mod.en.met)
```

2.  `plot` the model to look for deviations from modeling assumptions

```{r, eval = FALSE}
  plot(mod.en.met)
```

Select one or more additional predictors to add to your model and 
repeat steps 1-3. Is this model significantly better than the model 
with *metro* as the only predictor?

```{r, eval = FALSE}
  states.en.met.pop.wst <- subset(states, select = c("energy", "metro", "pop", "waste"))
  summary(states.en.met.pop.wst)
  plot(states.en.met.pop.wst)
  cor(states.en.met.pop.wst, use = "pairwise")
  mod.en.met.pop.waste <- lm(energy ~ metro + pop + waste, data = states)
  summary(mod.en.met.pop.waste)
  anova(mod.en.met, mod.en.met.pop.waste)
```

Exercise 1: prototype
---------------------

Use the states data set.

1.  Add on to the regression equation that you created in exercise 1 
by generating an interaction term and testing the interaction.

```{r, eval = FALSE}
  mod.en.metro.by.waste <- lm(energy ~ metro * waste, data = states)
```

2.  Try adding a region to the model. Are there significant 
differences across the four regions?

```{r, eval = FALSE}
  mod.en.region <- lm(energy ~ metro * waste + region, data = states)
  anova(mod.en.region)
```

Exercise 2 prototype
--------------------

Use the NH11 data set that we loaded earlier. Note that the 
data is not perfectly clean and ready to be modeled. You will
need to clean up at least some of the variables before fitting
the model.

1.  Use glm to conduct a logistic regression to predict ever 
worked (everwrk) using age (age_p) and marital status (r_maritl).

```{r, eval = FALSE}
  nh11.wrk.age.mar <- subset(NH11, select = c("everwrk", "age_p", "r_maritl"))
  summary(nh11.wrk.age.mar)
  NH11 <- transform(NH11,
                    everwrk = factor(everwrk,
                        levels = c("1 Yes", "2 No")),
                    r_maritl = droplevels(r_maritl))

  mod.wk.age.mar <- glm(everwrk ~ age_p + r_maritl, data = NH11,
                        family = "binomial")

  summary(mod.wk.age.mar)
```

2.  Predict the probability of working for each level of marital status.

```{r, eval = FALSE}
  library(effects)
  data.frame(Effect("r_maritl", mod.wk.age.mar))
```

Exercise 3 prototype {.smaller}
--------------------

Use the dataset, bh1996:

```{r, eval = FALSE}
  data(bh1996, package="multilevel")
```

From the data documentation:

> Variables are Cohesion (COHES), Leadership Climate (LEAD), 
> Well-Being (WBEING) and Work Hours (HRS). Each of these variables 
> has two variants - a group mean version that replicates each group
> mean for every individual, and a within-group version where the 
> group mean is subtracted from each individual response. The group 
> mean version is designated with a G. (e.g., G.HRS), and the 
> within-group version is designated with a W. (e.g., W.HRS).

Note that the group identifier is named "GRP".

1.  Create a null model predicting wellbeing ("WBEING")

```{r, eval = FALSE}
  library(lme4)
  mod.grp0 <- lmer(WBEING ~ 1 + (1 | GRP), data = bh1996)
  summary(mod.grp0)
```

2.  Run a second multi-level model that adds two 
individual-level predictors, average number of hours 
worked ("HRS") and leadership skills ("LEAD") to the 
model and interpret your output.

```{r, eval = FALSE}
  mod.grp1 <- lmer(WBEING ~ HRS + LEAD + (1 | GRP), data = bh1996)
  summary(mod.grp1)
```

3.  Now, add a random effect of average number of hours 
worked ("HRS") to the model and interpret your output. 
Test the significance of this random term.

```{r, eval = FALSE}
  mod.grp2 <- lmer(WBEING ~ HRS + LEAD + (1 + HRS | GRP), data = bh1996)
  anova(mod.grp1, mod.grp2)
```
