---
title: "R graphics with ggplot2 Exercises"
always_allow_html: yes
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
jupyter:
  jupytext_format_version: '1.0'
  jupytext_formats: ipynb,Rmd:rmarkdown,R
  kernelspec:
    display_name: R
    language: R
    name: ir
  language_info:
    codemirror_mode: r
    file_extension: .r
    mimetype: text/x-r-source
    name: R
    pygments_lexer: r
    version: 3.5.1
  toc:
    base_numbering: 1
    nav_menu: {}
    number_sections: true
    sideBar: true
    skip_h1_title: false
    title_cell: Table of Contents
    title_sidebar: Contents
    toc_cell: false
    toc_position: {}
    toc_section_display: true
    toc_window_display: true
---

## Exercise I

The data for the exercises is available in the `dataSets/EconomistData.csv` file. Read it in with

```{r results="hide", message=FALSE}
library(readr)
dat <- read_csv("dataSets/EconomistData.csv")
```

Original sources for these data are <http://www.transparency.org/content/download/64476/1031428> <http://hdrstats.undp.org/en/indicators/display_cf_xls_indicator.cfm?indicator_id=103106&lang=en>

These data consist of *Human Development Index* and *Corruption Perception Index* scores for several countries.

1.  Create a scatter plot with CPI on the x axis and HDI on the y
    axis.
```{r}

```
2.  Color the points blue.
```{r}

```
3.  Map the color of the the points to Region.
```{r}

```
4.  Make the points bigger by setting size to 2
```{r}

```
5.  Map the size of the points to HDI.Rank


## Exercise II

1.  Re-create a scatter plot with CPI on the x axis and HDI on the y
    axis (as you did in the previous exercise).
```{r}

```
2.  Overlay a smoothing line on top of the scatter plot using
    `geom_smooth`.
```{r}

```
3.  Overlay a smoothing line on top of the scatter plot using
    `geom_smooth`, but use a linear model for the predictions. Hint:
    see `?stat_smooth`.
```{r}

```
4.  Overlay a smoothing line on top of the scatter plot using
    `geom_line`. Hint: change the statistical transformation.
```{r}

```
5.  BONUS: Overlay a smoothing line on top of the scatter plot using
    the default *loess* method, but make it less smooth. Hint: see
    `?loess`.
```{r}

```

## Exercise III

1.  Create a scatter plot with CPI on the x axis and HDI on the y
    axis. Color the points to indicate region.
```{r}

```
2.  Modify the x, y, and color scales so that they have more
    easily-understood names (e.g., spell out "Human development Index"
    instead of "HDI").
```{r}

```
3.  Modify the color scale to use specific values of your choosing.
    Hint: see `?scale_color_manual`.

## Challenge: Recreate This `Economist` Graph

[images/Economist1.pdf](images/Economist1.pdf)

Graph source: <http://www.economist.com/node/21541178>

Building off of the graphics you created in the previous exercises,
put the finishing touches to make it as close as possible to the
original economist graph.
