---
title: "Exercises"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Exercise 0

The purpose of this exercise is to give you an opportunity to explore
the interface provided by RStudio (or whichever GUI you've decided to
use). You may not know how to do these things; that's fine! This is an
opportunity to figure it out.

Also keep in mind that we are living in a golden age of tab completion.
If you don't know the name of an R function, try guessing the first two
or three letters and pressing TAB. If you guessed correctly the function
you are looking for should appear in a pop up!

1.  Try to get R to add 2 plus 2.

2.  Try to calculate the square root of 10.

3.  R includes extensive documentation, including a file named "An
    introduction to R". Try to find this help file.


## Exercise 1: Reading data

The purpose of this exercise is to practice reading data into R.

1.  Open the help page for the `read_csv` function. How can you limit
    the number of rows to be read in?
```{r}
##
```
2.  Read just the first 10 rows of
`http://tutorials.iq.harvard.edu/data/babyNames.csv`

3.  Once you have successfully read in the first 10 rows, read the whole
    file, assigning the result to the name `baby.names`.


## Exercise 2: Data Extraction

Read in the "babyNames.csv" file if you have not already done so,
assigning the result to `baby.names`. The file is located at "http://tutorials.iq.harvard.edu/data/babyNames.csv"

1.  Extract data for the name "ashley".  

2.  Arrange the "ashley" data you produced above by `Count`. In which year was Ashley most popular?

3.  Restrict the previous extraction to include only years between 2000 and 2004.

## Exercise 3: Data manipulation

Read in the `http://tutorials.iq.harvard.edu/data/babyNames.csv` file if you have not already done so, assigning the result to
`baby.names`.

1.  If you look at `unique(baby.names$Sex)` you'll notice that some
    records indicate Male with `"M"`, while other records use
    `"Male"`. Correct this by replacing `"M"` with `"Male"`.

2.  Use `mutate` and `group_by` to create a column named "Proportion". Note that this step is exactly the same as shown in the demonstration. You can copy/paste that code if you like.

3.  Create a column named "Popular" containing `TRUE` in rows where "Proportion" is greater than 0.02 and a `FALSE` otherwise.

4.  Filter the baby names data to display only the popular names.

## Exercise 4: Group by and summarize

1.  Calculate the total number of children born.

2.  Calculate the number of boys and girls born each year. Assign the result
    to the name `births.by.year`.

3.  How many children were born in 2004?


## Exercise 5: Popularity of the most popular names
----------

In this exercise we will plot trends in the proportion of boys and girls given the most popular name each year.

1.  Filter the baby.names data, retaining only the most popular girl
    and boy names for each year.

2.  Plot proportion over time to see changes in the proportion of
    who chose the most popular name of the year.