---
title: "Introduction to R workshop notes"
always_allow_html: yes
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
jupyter:
  jupytext_format_version: '1.0'
  jupytext_formats: ipynb
  kernelspec:
    display_name: R
    language: R
    name: ir
  language_info:
    codemirror_mode: r
    file_extension: .r
    mimetype: text/x-r-source
    name: R
    pygments_lexer: r
    version: 3.5.1
---

```{r echo=FALSE, results='hide'}
options(max.print = 100)
knitr::opts_chunk$set(message = FALSE)
```

Welcome
========

Materials and setup 
-------------------

**NOTE: skip this section if you are not running R locally** (e.g., if you are
running R in your browser using a remote Jupyter server)

You should have R installed --if not:

-   Download and install R from http://cran.r-project.org
-   Download and install RStudio from https://www.rstudio.com/products/rstudio/download/#download

Notes and examples for this workshop are available at 
[](http://tutorials.iq.harvard.edu/R/Rintro/Rintro.html)

Start RStudio create a new project:
-   On Windows click the start button and search for rstudio. On Mac
    RStudio will be in your applications folder.
-   In Rstudio go to `File -> New Project`.
-   Choose `New Directory` and `New Project`.
-   Choose a name and location for your new project directory.

Workshop goals and approach
---------------------------

In this workshop you will

-  learn R language basics and common idioms,
-  learn about the R package ecosystem,
-  practice reading files and manipulating data in R

A more general goal is to get you comfortable with R so that it seems less scary and mystifying than it perhaps does now. Note that this is by no means a complete or thorough introduction to R! It's just enough to get you started.

This workshop is relatively informal, example-oriented, and hands-on. We won't spend much time examining language features in detail. Instead we will work through an example, and learn some things about the R language and environment along the way.

As an example project we will analyze the popularity of baby names in the US from 1960 through 2017. Among the questions we will use R to answer are:

-  How many children were named "Jack" or "Jill" in 1996?
-  In which year was the name "Ashley" the most popular?
-  What are the most popular names overall? For girls? For Boys?
-  How many children were born each year?

What is R?
----------

R is a *programming language designed for statistical computing*.
Notable characteristics include:

-   Vast capabilities, wide range of statistical and graphical
    techniques
-   Very popular in academia, growing popularity in business:
    <http://r4stats.com/articles/popularity/>
-   Written primarily by statisticians
-   FREE (no monetary cost and open source)
-   Excellent community support: mailing list, blogs, tutorials
-   Easy to extend by writing new functions

Whatever you're trying to do, you're probably not the first to try doing
it R. Chances are good that someone has already written a package for
that.

Graphical User Interfaces (GUIs)
================================

There are many different ways you can interact with R. See the
[Data Science Tools workshop notes](http://tutorials.iq.harvard.edu/DataScienceTools/DataScienceTools.html) 
for details.

For this workshop I encourage you to use RStudio; it is a good
R-specific IDE that mostly just works.

Launch RStudio (skip if not using Rstudio)
------------------------------------------
**Note: skip this section if you are not using Rstudio (e.g., if you are running
these examples in a Jupyter notebook).**

-   Start the RStudio program
-   In RStudio, go to **File =&gt; New File =&gt R Script**

The window in the upper-left is your R script. This is where you will
write instructions for R to carry out.

The window in the lower-left is the R console. This is where results
will be displayed.

Exercise 0
----------

The purpose of this exercise is to give you an opportunity to explore
the interface provided by RStudio (or whichever GUI you've decided to
use). You may not know how to do these things; that's fine! This is an
opportunity to figure it out.

Also keep in mind that we are living in a golden age of tab completion.
If you don't know the name of an R function, try guessing the first two
or three letters and pressing TAB. If you guessed correctly the function
you are looking for should appear in a pop up!

-------------------------------------------

1.  Try to get R to add 2 plus 2.
```{r}
##
```
2.  Try to calculate the square root of 10.
```{r}
##
```
3.  R includes extensive documentation, including a file named "An
    introduction to R". Try to find this help file.


Exercise 0 solution<span class="tag" data-tag-name="prototype"></span>
----------------------------------------------------------------------


```{r, purl = FALSE}
## 1. 2 plus 2
2 + 2
## or
sum(2, 2)

## 2. square root of 10:
sqrt(10)
## or
10^(1/2)


## 3. Find "An Introduction to R".

## Go to the main help page by running 'help.start() or using the GUI
## menu, find and click on the link to "An Introduction to R".

```


R basics
========


Function calls
---------

The general form for calling R functions is

```{r eval=FALSE}
## FunctionName(arg.1 = value.1, arg.2 = value.2, ..., arg.n - value.n)
```

Arguments can be matched by name; unnamed arguments will be matched by position.

Assignment
----------

Values can be assigned names and used in subsequent operations

-   The `<-` operator (less than followed by a dash) is used to save
    values
-   The name on the left gets the value on the right.

```{r}
sqrt(10) ## calculate square root of 10; result is not stored anywhere
x <- sqrt(10) # assign result to a variable named x
```

Asking R for help
---------------------

You can ask R for help using the `help` function, or the `?` shortcut.

```{r}
help(help)
```

The `help` function can be used to look up the documentation for a function, or
to look up the documentation to a package. We can learn how to use the `stats`
package by reading its documentation like this:

```{r, eval=FALSE}
help(package = "stats")
```


Getting data into R
===================

R has data reading functionality built-in -- see e.g.,
`help(read.table)`. However, faster and more robust tools are
available, and so to make things easier on ourselves we will use a
*contributed package* called `readr` instead. This requires that we
learn a little bit about packages in R.

Installing and using R packages
----------------------------------------------------

A large number of contributed packages are available. If you are
looking for a package for a specific task,
https://cran.r-project.org/web/views/ and https://r-pkg.org are good
places to start.

You can install a package in R using the `install.packages()`
function. Once a package is installed you may use the `library`
function to attach it so that it can be used.

```{r}
## install.packages("readr")
library(readr)
```

Readers for common file types
-----------------------------

In order to read data from a file, you have to know what kind of file
it is. The table below lists functions that can import data from
common plain-text formats.

| Data Type                 | Function        |
| ------------------------- | --------------- |
| comma separated           | `read_csv()`    |
| tab separated             | `read_delim()`  |
| other delimited formats   | `read_table()`  |
| fixed width               | `read_fwf()`    |

**Note** You may be confused by the existence of similar functions,
e.g., `read.csv` and `read.delim`. These are legacy functions that
tend to be slower and less robust than the `readr` functions. One way
to tell them apart is that the faster more robust versions use
underscores in their names (e.g., `read_csv`) while the older
functions us dots (e.g., `read.csv`). My advice is to use the more
robust newer versions, i.e., the ones with underscores.

Baby names data
---------------

The examples in this workshop use US baby names data retrieved from
https://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data
A cleaned and merged version of these data is available at
`http://tutorials.iq.harvard.edu/data/babyNames.csv`.


Exercise 1: Reading the baby names data
---------------------------------------

The first step of our analysis of baby names is to read the data into R. 

1. Start by opening the help page for the `read_csv` function so you can refer
to the documentation if needed.

2. Read the baby names using the `read_csv` function and assign the result 
with the name `baby.names`.

3. BONUS (optional): Using 
https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf 
as a guide, save the `baby.names` data as a Stata data set `babynames.dta` 
and as an R data set `babynames.rds`.

Exercise 1 solution<span class="tag" data-tag-name="prototype"></span>
----------------------------------------------------------------------

```{r, results = FALSE, message = FALSE, purl = FALSE}
## read ?read_csv

baby.names <- read_csv("http://tutorials.iq.harvard.edu/data/babyNames.csv")
```

Popularity of your name
=======================

In this section we will pull out specific names and examine changes in 
their popularity over time. 

The `baby.names` object we created in the last exercise is a `data.frame`.
There are many other data structures in R, but for now we'll focus on 
working with `data.frames`. 

R has decent data manipulation tools built-in -- see e.g.,
`help(Extract)`. However, these tools are powerful and complex and
often overwhelm beginners. To make things easier on ourselves we will
use a *contributed package* called `dplyr` instead.

```{r}
## install.packages("dplyr")
library(dplyr)
```

Filtering and arranging data
----------------------------

One way to find the year in which your name was the most popular
is to filter out just the rows corresponding to your name, and 
then arrange (sort) by Count. 

To demonstrate these techniques we'll try to determine whether 
"Alex"" or "Jim" was more popular in 1992. We start by filtering the
data so that we keep only rows where Year is equal to `1992` and Name is
either "Alex" or "Mark".

```{r}
am <- filter(baby.names, 
             Year == 1992 & (Name == "Alex" | Name == "Mark"))
am
```


Notice that we can we can combine conditons using `&` (AND) 
and `|` (OR). 

In this case it's pretty easy to see that "Mark" is more popular, 
but to make it even easier we can arrange the data so that the 
most popular name is listed first.

```{r}
arrange(am, Count)
```

```{r}
arrange(am, desc(Count))
```


Other logical operators
-----------------------

In the previous example we used `==` to filter rows. Other relational
and logical operators are listed below.

 | Operator  | Meaning                   | 
 | ----------| --------------------------| 
 | `==`      | equal to                  | 
 | `!=`      | not equal to              | 
 | `>`       | greater than              | 
 | `>=`      | greater than or equal to  | 
 | `<`       | less than                 | 
 | `<=`      | less than or equal to     | 
 | `%in%`    | contained in              | 

These operators may be combined with `&` (and) or `|` (or).


Exercise 2: Peak popularity of your name
-----------------------------

Read in the "babyNames.csv" file if you have not already done so,
assigning the result to `baby.names`. The file is located at 
`"http://tutorials.iq.harvard.edu/data/babyNames.csv"`

1.  Use `filter` to extract data for your name (or another name of your choice).  
```{r}
##
```

2.  Arrange the data you produced in step 1 above by `Count`. 
    In which year was the name most popular?
```{r}
##
```    

3.  BONUS (optional): Identify the most popular boys name in 1999.
```{r}
##
```


Exercise 2 solution<span class="tag" data-tag-name="prototype"></span>
----------------------------------------------------------------------

```{r, purl = FALSE}
# 1.  Use `filter` to extract data for your name (or another name of your choice).  

george <- filter(baby.names, Sex == "Boys" & Name == "George")

# 2.  Arrange the data you produced in step 1 above by `Count`. 
#     In which year was the name most popular?

arrange(george, desc(Count))
 
# 3.  BONUS (optional): Identify the most popular boys name in 1999.

arrange(filter(baby.names, Year == 1999 & Sex == "Boys"),
        desc(Count))

```


Finding the most popular names
==============================

Our next goal is to find out which names have been the most popular.

Computing better measures of popularity
---------------------------------------

So far we've used `Count` as a measure of popularity. A better
approach is to use proportion or rank to avoid confounding 
popularity with the number of babies born in a given year. 

The `mutate` function makes it easy to add or modify the columns 
of a `data.frame`. For example, we can use it compute the 
log of the number of boys and girls given each name in each year:

```{r}
baby.names <- mutate(baby.names, logCount = Count/1000)
baby.names
```

Operating by group
------------------

Because of the nested nature of out data, we want to compute rank 
or proportion  within each `Sex` `X` `Year` group. The `dplyr` 
package makes this relatively easy.


```{r}
baby.names <- mutate(group_by(baby.names, Year, Sex),
                     Rank = rank(Count))
```

Note that the data remains grouped until you change the groups by
running `group_by` again or remove grouping information with
`ungroup`.

Exercise 3: Most popular names
-----------------------------

In this exercise your goal is to identify the most popular names for each year.

1.  Use `mutate` and `group_by` to create a column named "Proportion"
    where `Proportion = Count/sum(Count)` for each `Year X Sex` group.
```{r}
##
```

2.  Use `mutate` and `group_by` to create a column named "Rank" where 
    `Rank = rank(-Count)` for each `Year X Sex` group. 

```{r}
##
```

3.  Filter the baby names data to display only the most popular name 
    for each `Year X Sex` group.
```{r}
##
```

4. BONUS (optional): Which names are the most popular for both boys and girls?
```{r}
##
```
Exercise 3 solution<span class="tag" data-tag-name="prototype"></span>
----------------------------------------------------------------------

```{r, purl = FALSE}
## 1.  Use `mutate` and `group_by` to create a column named "Proportion"
##     where `Proportion = Count/sum(Count)` for each `Year X Sex` group.

baby.names <- mutate(group_by(baby.names, Year, Sex),
                     Proportion = Count/sum(Count))

## 2.  Use `mutate` and `group_by` to create a column named "Rank" where 
##     `Rank = rank(-Count)` for each `Year X Sex` group.

baby.names <- mutate(group_by(baby.names, Year, Sex),
                     Rank = rank(-Count))

## 3.  Filter the baby names data to display only the most popular name 
##     for each `Year X Sex` group.

filter(baby.names, Rank == 1)

## 4. BONUS (optional): Which names are the most popular for both boys 
##    and girls?

girls.and.boys <- inner_join(filter(baby.names, Sex == "Boys"), 
                             filter(baby.names, Sex == "Girls"),
                             by = c("Year", "Name"))

girls.and.boys <- mutate(girls.and.boys,
                         Product = Count.x * Count.y,
                         Rank = rank(-Product))

filter(girls.and.boys, Rank == 1)
```

Percent choosing one of the top 10 names
========================================

You may have noticed that the percentage of babies given the most 
popular name of the year appears to have decreases over time. We can
compute a more robust measure of the popularity of the most popular
name by calculating the number of babies given one of the top 10 girl
or boy names of the year.

In order to compute this measure we need to operate within goups, as
we did using `mutate` above, but this time we need to collapse each
group into a single summary statistic. We can achive this using the
`summarize` function. For example, we can calculate the number of
babies born each year:

```{r}
bn.by.year <- summarize(group_by(baby.names, Year),
                       Total = sum(Count))
bn.by.year
```


Exercise 4: Proportion choosing a top-10 name
----------------------------------------------

1.  Calculate the proportion of girls given one of 
    the top 10 names each year. Does the proportion
    of babies given one of the top 10 names increase or
    decrease over time?
```{r}
```

2.  Calculate the proportion of boys given one of 
    the top 10 names each year. Does the proportion
    of babies given one of the top 10 names increase or
    decrease over time?
```{r}
##
```

3.  Bonus (optional): Which name has increased (from 1960 to 2017) 
                      in popularity the most?
```{r}
##
```
Exercise 4 solution<span class="tag" data-tag-name="prototype"></span>
----------------------------------------------------------------------

```{r, purl = FALSE}
## 1.  Calculate the proportion of girls given one of 
##     the top 10 names each year. Does the proportion
##     of babies given one of the top 10 names increase or
##     decrease over time?
top10girls <- summarize(group_by(filter(baby.names, Sex == "Girls", Rank <= 10),
                            Year, Sex),
                   Proportion = sum(Proportion))
top10girls

## 2.  Calculate the proportion of boys given one of 
##     the top 10 names each year. Does the proportion
##     of babies given one of the top 10 names increase or
##     decrease over time?
top10boys <- summarize(group_by(filter(baby.names, Sex == "Boys", Rank <= 10),
                            Year, Sex),
                   Proportion = sum(Proportion))
top10boys

## 3.  Bonus (optional): Which name has increased (from 1960 to 2017) 
##                       in popularity the most?

inboth <- inner_join(filter(baby.names, Year == 1960),
                     filter(baby.names, Year == 2017),
                     by = c("Name", "Sex"))

inboth <- mutate(group_by(inboth, Sex), 
                 Change = Proportion.y - Proportion.x,
                 ChangeRank = rank(-Change))

filter(inboth, ChangeRank == 1)

```

Plotting baby name trends over time
===================================

It can be difficult to spot trends when looking at summary tables.
Plotting the data makes it easier to identify interesting patterns.
Earlier we calculated the proportion of babies given one of the top
10 names each year. Plotting these data will allow us to more easily
spot changes in this index over time.

R has decent plotting tools built-in -- see e.g., `help(plot)`.
However, To make things easier on ourselves we will use a *contributed
package* called `ggplot2` instead.

```{r}
## install.packages("ggplot2")
library(ggplot2)
```

For quick and simple plots we can use the `qplot` function. For example,
we can plot the number of babies born over time like this:

```{r}
bn.by.year <- summarize(group_by(baby.names, Year),
                        Total = sum(Count))
qplot(x = Year, y = Total,
     geom = "line",
     data = bn.by.year) +
  ylim(0, max(bn.by.year$Total))
```

Splitting these trends out by Sex is very easy:

```{r}
bn.by.year.sex <- summarize(group_by(baby.names, Year, Sex),
                        Total = sum(Count))
qplot(x = Year, y = Total, color = Sex,
     geom = "line",
     data = bn.by.year.sex) +
     ylim(0, max(bn.by.year.sex$Total))
```

Exercise 5: Popularity of the most popular names
------------------------------------------------

In this exercise we will plot trends in the proportion of boys and girls given one of the 10 most popular names each year.

1.  Filter the baby.names data, retaining only the 10 most popular girl
    and boy names for each year.
```{r}
##
```

2.  Summarize the data produced in step one to calculate the total
    Proportion of boys and girls given one of the top 10 names
    each year.
```{r}
##
```

3.  Plot the data produced in step 2, with year on the x-axis
    and total proportion on the y axis. Color by sex.
```{r}
##
```

Exercise 5 solution<span class="tag" data-tag-name="prototype"></span>
----------------------------------------------------------------------

```{r, purl = FALSE}
## 1.  Filter the baby.names data, retaining only the 10 most 
##     popular girl and boy names for each year.

most.popular <- filter(group_by(baby.names, Year, Sex),
                       Rank <= 10)

## 2.  Summarize the data produced in step one to calculate the total
##     Proportion of boys and girls given one of the top 10 names
##     each year.

top10 <- summarize(group_by(most.popular, Year, Sex),
                   TotalProportion = sum(Proportion))
 
## 3.  Plot the data produced in step 2, with year on the x-axis
##     and total proportion on the y axis. Color by sex.

qplot(x = Year, y = TotalProportion, color = Sex,
      data = top10,
      geom = "line")
```


Saving our Work
==============

Now that we have made some changes to our data set, we might want to
save those changes to a file.

Saving individual datasets
--------------------------

```{r}
# write data to a .csv file
write_csv(baby.names, "babyNames.csv")

# write data to an R file
write_rds(baby.names, "babyNames.rds")
```

Saving and loading R workspaces
-------------------------------

In addition to importing individual datasets, R can save and load entire
workspaces

```{r}
ls() # list objects in our workspace
save.image(file="myWorkspace.RData") # save workspace 
rm(list=ls()) # remove all objects from our workspace 
ls() # list stored objects to make sure they are deleted

## Load the "myWorkspace.RData" file and check that it is restored
load("myWorkspace.RData") # load myWorkspace.RData
ls() # list objects
```


Wrap-up
=======

Help us make this workshop better!
----------------------------------

Please take a moment to fill out a very short feedback form. These
workshops exist for you -- tell us what you need!
<http://tinyurl.com/R-intro-feedback>

Additional resources
--------------------

-   IQSS workshops:
    <http://projects.iq.harvard.edu/rtc/filter_by/workshops>
-   IQSS statistical consulting: <http://dss.iq.harvard.edu>
-   Software (all free!):
    -   R and R package download: <http://cran.r-project.org>
    -   Rstudio download: <http://rstudio.org>
    -   ESS (emacs R package): <http://ess.r-project.org/>
-   Online tutorials
    -   <http://www.codeschool.com/courses/try-r>
    -   <http://www.datacamp.org>
    -   <http://swirlstats.com/>
    -   <http://r4ds.had.co.nz/>
-   Getting help:
    -   Documentation and tutorials:
        <http://cran.r-project.org/other-docs.html>
    -   Recommended R packages by topic:
        <http://cran.r-project.org/web/views/>
    -   Mailing list: <https://stat.ethz.ch/mailman/listinfo/r-help>
    -   StackOverflow: <http://stackoverflow.com/questions/tagged/r>
-   Coming from...
    Stata
    :   <http://www.princeton.edu/~otorres/RStata.pdf>
    SAS/SPSS
    :   <http://www.et.bs.ehu.es/~etptupaf/pub/R/RforSAS&SPSSusers.pdf>
   matlab
   :   <http://www.math.umaine.edu/~hiebeler/comp/matlabR.pdf>
    Python
    :   <http://mathesaurus.sourceforge.net/matlab-python-xref.pdf>
