---
title: Introduction To Programming In R
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---

Workshop overview and materials
===============================

Workshop description
--------------------

This is an intermediate/advanced R course appropriate for those with basic
knowledge of R. It is intended for those already comfortable with using R for
data analysis who wish to move on to writing their own functions.

Prerequisite: basic familiarity with R, such as acquired from an introductory R
workshop.

To the extent possible this workshop uses real-world examples. That is to say
that concepts are introduced as they are needed for a realistic analysis task.
In the course of working through a realistic project we will lean about regular
expressions, iteration, functions, control flow and more.

This workshop uses
the [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html)
package which provides more consistent file IO
([readr](https://cran.r-project.org/web/packages/readr/index.html)), data
manipulation
([dplyr](https://cran.r-project.org/web/packages/dplyr/index.html),
[tidyr](https://cran.r-project.org/web/packages/tidyr/index.html)) and
functional programming
([purrr](https://cran.r-project.org/web/packages/purrr/index.html)) tools for R.

Materials and setup
-------------------

**NOTE: skip this section if you are running these examples in your web browser at tutorials-live.iq.harvard.edu**

Everyone should have R installed -- if not:

-   Open a web browser and go to <http://cran.r-project.org> and download and
    install it
-   Also helpful to install RStudio (download from <http://rstudio.com>)

Materials for this workshop consists of notes and example code.

-   Download materials from
    <http://tutorials.iq.harvard.edu/R/RProgramming.zip>
-   Extract the zip file containing the materials to your desktop

Example project overview
------------------------

It is common for data to be made available on a website somewhere, either by a
government agency, research group, or other organizations and entities. Often
the data you want is spread over many files, and retrieving it all one file at a
time is tedious and time consuming. Such is the case with the baby names data we
will be using today.

The UK [Office for National Statistics](https://www.ons.gov.uk) provides yearly
data on the most popular baby names going back to 1996. The data is provided
separately for boys and girls. These data have been cleaned up and copied
to <example_data/baby_names/>.

Locating data in a nested directory structure (string manipulation)
=============================================================

Our first goal is to locate all the all the .csv files from containing data from
England and Wales. As a first step, we'll recursively list all the files in the `example_data/baby_names` directory.


```{r}
data_dir_files <- list.files("example_data/baby_names",
                             recursive = TRUE,
                             full.names = TRUE)

head(data_dir_files)
```

Our next task is to filter this directory listing so that we have only the .csv
files from England and Wales.

String manipulation packages
-----------------------------

In order extract the data files from the directory listings we will use functions for
manipulating strings[^1]. What functions or packages should we use? Here are
some tools to help us decide.

Task views
:   <https://cran.r-project.org/web/views/NaturalLanguageProcessing.html>

R package search
:   <http://www.r-pkg.org/search.html?q=string>

Web search
:   <https://www.google.com/search?q=R+string+manipulation&ie=utf-8&oe=utf-8>

Base R provides some string manipulation capabilities (see `?regex`, `?sub` and
`?grep`), but I recommend either
the [stringr](https://cran.r-project.org/web/packages/stringr/index.html)
or [stringi](https://cran.r-project.org/web/packages/stringi/index.html)
packages. The *stringr* package is more user-friendly, and the *stringi* package
is more powerful and complex. Let's use the friendlier one. While we're
attaching packages we'll also attach
the [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html)
package which provides more modern versions of many basic functions in R.


String manipulation with the stringr package
--------------------------------------------

Regular expressions are useful in general (not just in R!) and it is a good idea
to be familiar with at least the basics.

In regulars expression `^`, `.`, `*`, `+`, `$` and `\` are special characters
with the following meanings:

\^
:   matches the beginning of the string

.
:   matches any character

\*
:   repeats the previous match zero or more times

+
:   repeats the previous match one or more times

\$
:   matches the end of the string

\[\]
:   specifies ranges of characters: \[a-z\] matches lower case letters

\
:   escapes special meaning: '.' means "anything", '.' means "."

Here's how it works in R using the stringr package. Note that to create
a backslash you must escape it with anouth backslash.

```{r}
## make up some example data
user.info <- c("Dexter Bacon dbacon@gmail.com 32",
               "Angelica Sampson not available 28",
               "Roberta Modela roberta.modela@harvard.edu 26"
               )
## regex that matches emails (simplified, not realistic)
##                                               
email.regex <- "([a-z0-9]+@[a-z0-9]+\\.[a-z]+)"
```

Which users have an email address?
```{r}
str_detect(user.info, email.regex)
```

Keep only users with an email address
```{r}
str_subset(user.info, email.regex)
```

Extract the email addresses
```{r}
str_extract(user.info, email.regex)
```

Replace email addresses with an `mailto` link
```{r}
str_replace(user.info, email.regex, "<a href='mailto:\\1'>\\1</a>")
```

If you have not been introduced to regular expressions yet a nice interactive
regex tester is available at <http://www.regexr.com/> and an interactive
tutorial is available at <http://www.regexone.com/>.

Exercise 1: string manipulation and regular expressions
-------------------------------------------------------

Our job now is to match the file names using regular expressions. To get
started lets copy an example string into the interactive regex tester at
<http://www.regexr.com/> and work with it until we find a regular
expression that works.

1.  Open <http://www.regexr.com/> in your web browser and paste in this
    string into the "Text" box (in the lower right-hand section of the
    page):

```{.example}
example_data/baby_names/EW.1
example_data/baby_names/EW/boys_1996.csv
example_data/baby_names/EW/boys_1997.csv
example_data/baby_names/EW/boys_1998.csv
example_data/baby_names/EW/boys_1999.csv
example_data/baby_names/EW/girls_1996.csv
example_data/baby_names/EW/girls_1997.csv
example_data/baby_names/EW/girls_1998.csv
example_data/baby_names/EW/girls_1999.csv
example_data/baby_names/EW/index.html
example_data/baby_names/EW/index.html?C=D;O=A
example_data/baby_names/EW/index.html?C=D;O=D
example_data/baby_names/EW/index.html?C=M;O=A
example_data/baby_names/EW/index.html?C=M;O=D
example_data/baby_names/EW/index.html?C=N;O=A
example_data/baby_names/index.html?C=S;O=A
example_data/baby_names/index.html?C=S;O=D
example_data/baby_names/US/AK.csv
example_data/baby_names/US/AL.csv
example_data/baby_names/US/AR.csv
example_data/baby_names/US/AZ.csv
example_data/baby_names/US/CA.csv
example_data/baby_names/US/CO.csv
example_data/baby_names/US/CT.csv
```

2.  Find a regular expression that matches lines corresponding to data files for
    girls from England and Wales. Make sure that you have _not_ matched any
    other lines.
3.  Assign the regular expression you found to the name
    'girl.file.regex' in R. *Replace any backslashes with a double
    backslash*.
```{r}

```
4.  Extract the England and Wales data file names from the directory listing and
    assign the values to the name '`girl.file.names`'
```{r}

```
5.  Repeat steps 2-4 for the boys data.
\

\

\

\

\


Exercise 1 prototype
--------------------

```{r, results = 'hide', purl = FALSE}
## 1. Make sure you replace whatever example text was presented in the regexr.com site.

## 2.  
## 3.  Assign the regular expression you found to the name
##    'girl.file.regex' in R. *Replace any backslashes with a double
##     backslash*.

girl.file.regex <- "EW.*girls.*\\.csv$"

## 4.  Extract the England and Wales data file names from the directory listing and
##     assign the values to the name '`girl.file.names`'
library(stringr)

girl.file.names <- str_subset(data_dir_files, girl.file.regex)

## 5.  Repeat steps 3-4 for the boys data.
boy.file.regex <- "EW.*boys.*\\.csv$"
boy.file.names <- str_subset(data_dir_files, boy.file.regex)
```

Reading all the files (iteration, functions)
============================================

Reading .csv files
------------------

we can read the first girls names data file like this:

```{r}
library(tidyverse)

girl.names.1 <- read_csv(girl.file.names[1], na = "")
head(girl.names.1)
```

Notice that we selected the first element of `girl.file.names` using `[`. This
is called *bracket extraction* and it is a very useful feature of the R
language.

Extracting and replacing vector elements
----------------------------------------

Elements of R objects can be extracted and replaced using bracket
notation. Bracket extraction comes in a few different flavors. We can
index atomic vectors in several different ways. Let's start by making
some example data to work with.

```{r}
example.int.1 <- c(10, 11, 12, 13, 14, 15)
str(example.int.1)
```

The *names* attribute can be used for indexing if it exists. The names
attribute and be extracted and set using the `names` function, like
this:

```{r}
names(example.int.1) # no names yet, lets add some
```

```{r}
names(example.int.1) <- c("a1", "a2", "b1", "b2", "c1", "c2")
names(example.int.1)
```

Indexing can be done by position.

```{r}
## extract by position
example.int.1[1]
example.int.1[c(1, 3, 5)]
```

If an object has names you can index by name.

```{r}
## extract by name
example.int.1[c("c2", "a1")]
```

Finally, objects can be indexed with a logical vector, extracting only the TRUE
elements.

```{r}
## logical extraction 
(one.names <- str_detect(names(example.int.1), "1"))

```

```{r}
example.int.1[one.names]
```

```{r}
example.int.1[example.int.1 > 12]
```

Replacement works by assigning a value to an extraction.

```{r}
example.int.2 <- example.int.1

## replace by position
example.int.2[1] <- 100

## replace by name
example.int.2["a2"] <- 200
example.int.2
```

```{r}
## logical replacement
(lt14 <- example.int.2 < 14)
```

```{r}
example.int.2[lt14] <- 0
example.int.2
```

```{r}
## "replace" non-existing element
example.int.2[c("z1", "2")] <- -10 

## compare lists to see the changes we made
example.int.1
example.int.2
```

Extracting and replacing list elements
--------------------------------------

List elements can be extracted and replaced in the same way as elements of
atomic vectors. In addition, `[[` can be used to extract or replace the contents
of a list element.

Back to the problem at hand: reading data into a list
-----------------------------------------------------

Using our knowledge of bracket extraction we could start reading in the data
files like this:

```{r}
boys <- list()
girls <- list()

boys[[1]] <- read_csv(boy.file.names[1], na = "")
boys[[2]] <- read_csv(boy.file.names[2], na = "")
## ...
girls[[1]] <- read_csv(girl.file.names[1], na = "")
girls[[2]] <- read_csv(girl.file.names[2], na = "")
## ...
```

Exercise 2: String manipulation, extraction and replacement
-----------------------------------------------------------

We saw in the previous example one way to start reading the baby name
data, using positional bracket extraction and replacement. In this
exercise we will improve on this method by doing the same thing using
named extraction and replacement. The first step is to extract the years
from `boy.file.names` and `girl.file.names` and assign then to the
`names` attribute of our `boys` and `girls` lists.

1.  Use the `list` function to create empty lists named `boys` and `girls`.
```{r}

```
2.  Write a regular expression that matches digits 0-9 repeated one or
    more times and use it to extract the years from `boy.file.names` and
    `girl.file.names` (use `str_extract`).
```{r}

```
3.  Use the assignment from of the `names` function to assign the years
    vectors from step one to the names of `boy.file.names` and
    `girl.file.names` respectively.
```{r}

```
4.  Extract the element named "2015" from `girl.file.names` and pass it
    as the argument to `read_csv`, assigning the result to a new element
    of the `girls` list named "2015". Repeat for elements "2014"
    and "2013".
```{r}

```
5.  Repeat step three using `boy.file.names` and the `boys` list.

\

\

\

\

\

\


Exercise 2 prototype
-------------------------


```{r, results = 'hide', purl = FALSE}
  ## 1. create lists to store the data
  boys <- list()
  girls <- list()

  ## 2. extract years and assign to names
  boy.years <- str_extract(boy.file.names, "[0-9]+")
  girl.years <- str_extract(girl.file.names, "[0-9]+")
  names(girl.file.names) <- girl.years
  names(boy.file.names) <- boy.years

  ## 3. read first three years of girls names
  girls[["2015"]] <- read_csv(girl.file.names["2015"], na = "")
  girls[["2014"]] <- read_csv(girl.file.names["2014"], na = "")
  girls[["2013"]] <- read_csv(girl.file.names["2013"], na = "")

  ## 4. read first three years of boys names
  boys[["2015"]] <- read_csv(boy.file.names["2015"], na = "")
  boys[["2014"]] <- read_csv(boy.file.names["2014"], na = "")
  boys[["2013"]] <- read_csv(boy.file.names["2013"], na = "")
 
  ## 5. examine the result
  str(boys, max.level = 2)
  str(girls, max.level = 2)
```

Iterating using the map function
--------------------------------

With a small number of files reading each one separately isn't too bad, but it
obviously doesn't scale well. To read all the files conveniently we instead want
to instruct R to iterate over the vector of URLs for us and download each one.
We can carry out this iteration in several ways, including using one of the
`map*` functions in the purrr package. Here is how it works.

```{r}
## make up some example data
list.1 <- list(a = sample(1:5, 20, replace = TRUE),
               b = c(NA, sample(1:10, 20, replace = TRUE)),
               c = sample(10:15, 20, replace = TRUE))
```

Calculate the mean of every entry
```{r}
library(purrr)

map.1 <- map(list.1, mean)
str(map.1)
```

Calculate the mean of every entry, passing `na.rm` argument.
```{r}
map.1 <- map(list.1, mean, na.rm = TRUE)
str(map.1)
```

Calculate the mean of every entry, returning a numberic vector instead of a
list.
```{r}
map.2 <- map_dbl(list.1, mean)
str(map.2)
```

Calculate the mean of every entry, returning a character vector.
```{r}
map.3 <- map_chr(list.1, mean)
```

Calculate summaries (map returns a list).
```{r}
map.4 <- map(list.1, summary)
str(map.4)
```

Writing your own functions
--------------------------

The `map*` functions are useful when you want to apply a function to a list or
vector of inputs and obtain the return values. This is very convenient when a
function already exists that does exactly what you want. In the examples above
we mapped `mean` and `summary` to the elements of a list. But what if there is
no existing function that does exactly what we want? Suppose that rather than
the set of statistics reported by the `summary` function we want to summarize
each element in the list by calculating the *length*, *mean*, and *standard
deviation*? In that case we will need to write a function that does what we
want. Fortunately, writing functions in R is easy.

```{r}
## define new function that returns the lenght, mean, and standard deviation
my.summary <- function(x) {
  n <- length(x)
  avg <- mean(x)
  std.dev <- sd(x)
  return(c(N = n, Mean = avg, Standard.Deviation = std.dev))
}

## test it out
my.summary(list.1[[1]])
```

Our new function can be used just like any other function in R. For example, we
can iterate over `list.1` and apply our new function to each element.

```{r}
map(list.1, my.summary)
```

Note that you can use the special `...` notation to pass named arguments
without needing to define them all. For example:

```{r}
## improve our function by allowing arbitrary arguments to be passed into the body
my.summary <- function(x, ...) {
  n <- length(x)
  avg <- mean(x, ...)
  std.dev <- sd(x, ...)
  return(c(N = n, Mean = avg, Standard.Deviation = std.dev))
}
```

Now we can try out the new version of our summary function. Calling it without
any additional arguments works fine for the first element of the list:
```{r}
my.summary(list.1[[1]])
```

However, it doesn't work so well on elements containing missing values:
```{r}
my.summary(list.1[[2]])
```

Even though our function does not have an na.rm argument we, can pass it to mean and sd via `...`

```{r}
my.summary(list.1[[2]], na.rm = TRUE) 
```

Often when writing functions you want to skip some part of the function body
under some conditions. For example, we might want omit missing values only if
they exist:

```{r}
## improve our function by removing missing values if they exist.
my.summary <- function(x, ...) {
  if(any(is.na(x))) {
    x <- na.omit(x)
  }
  n <- length(x)
  avg <- mean(x, ...)
  std.dev <- sd(x, ...)
  return(c(N = n, Mean = avg, Standard.Deviation = std.dev))
}
```

This is often useful for argument checking among other things.

```{r}
my.summary <- function(x, mean.only = FALSE, ...) {
  if(!is.numeric(x)) {
    stop("x is not numeric.")
  }
  if(any(is.na(x))) {
    x <- na.omit(x)
  }
  if(mean.only) {
    stats <- c(N = length(x), Mean = mean(x))
  } else {
    stats <- c(N = length(x), Mean = mean(x), Standard.Deviation = sd(x))
  }
  return(stats)
}
```

```{r}
map(list.1, my.summary)
```

```{r}
map(list.1, my.summary, mean.only = TRUE)
```

OK, now that we know how to write functions lets get back to the problem
at hand. We want to read each file in the `girl.file.names` and
`boy.file.names` vectors.

Exercise 3: Iteration, file IO, functions
-----------------------------------------

We know how to read csv files using `read_csv`. We know how to iterate
using `map`. All we need to do now is put the two things together.

1.  Use the `map` and `read_csv` functions to read all the girls data
    into an object named `girls.data`.
```{r}

```
2.  Do the same thing for the boys data (name the object `boys.data`).
```{r}

```
3.  Inspect the boys and girls data lists. How many elements do they
    have? What class are they?
```{r}

```
4.  Write a function that returns the `class`, `typeof`, and `length` of
    its argument. `map` this function to `girls.data` and `boys.data`.
```{r}

```

\

\

\

\

\

\ 

Exercise 3 prototype
-------------------------

```{r, results = 'hide', purl = FALSE}
## 1. read in the girls names data
girls.data <- map(girl.file.names, read_csv, na = "")

## 2. read in the boys names data
boys.data <- map(boy.file.names, read_csv, na = "")

## 3. inspect boys and girls names data
length(boys.data)
length(girls.data)
class(girls.data)
class(boys.data)

## 4. Inspect meta data for each set of names data
list.info <- function(x) return(c(class = class(x), mode = mode(x), n = length(x)))
unique(map(girls.data, list.info))
unique(map(boys.data, list.info))
```


Adding a Year column (data structures, indexing)
================================================

OK, at this point we have read in all the data. That is a good start! Lets take
a closer look at what we are working with.

```{r}
str(girls.data, max.level = 1)
```

OK, so we have a *list* with 20 elements. A list in R is a very flexible data
structure that can hold just about anything (including other lists). Our
`girls.data` is a list that happens to contain *data.frames*.

A data.frame in R is a special type of list in which each element is the same
length. Elements of data.frames are often *atomic vectors* ("logical",
"integer", "numeric (double)" etc. as discussed above), though elements of a
data.frame can be anything you can store in other lists, provided only that each
element is the same length.

There are many useful things we can do with data.frames: (But remember that this
list of methods is not all you can do with data.frames.)

```{r}
girls.data.1 <- girls.data[[1]]
class(girls.data.1)
```

```{r}
methods(class = "data.frame")
```

```{r}
list(dim = dim(girls.data.1),
     rows = nrow(girls.data.1),
     cols = ncol(girls.data.1),
     names = names(girls.data.1),
     summary = summary(girls.data.1))
```

Adding a year column using bracket extraction/replacement
---------------------------------------------------------

The data.frames in `girls.data` all have three elements (columns): Rank (a
number), Name (a character), and Count (a number). That is a problem, because we
need to have a column that records the year! Year was stored in the file names
rather than in the contents of the files, and so it appears as the names of the
list elements rather than as a column in each data.frame.

How can we insert the year information into each corresponding data.frame?
Recall that a data.frame **is** a list, so we can use bracket extraction just as
we would for other kinds of lists. That is, we can add a year column to the
data.frame in position 1 of `girls.data` like this:

```{r}
str(girls.data[[1]])
```

```{r}
girls.data[[1]][["Year"]] <- names(girls.data[1])
##        ^                                  ^ 
## double brackets to extract data.frame   single brackets to extract a list containing a data.frame
str(girls.data[[1]])
```

```{r}
girls.data[[1]][["Year"]] <- as.integer(names(girls.data[1]))
str(girls.data[[1]])
```

Now we want to repeat that for each element in the list. We previously learned
how to iterate using the `map` function, but that won't work here because we
need a function with two arguments: the list element to modify, and the year
value to insert. For that we can use the `map2` function, which works just like
`map` except that we iterate over two things with a function that takes two
arguments. It works like this:

```{r}
example.list.3 <- list(a = 1:3,
                       b = 4:6,
                       c = 7:9)
upper.name <- function(x, y) {
  y.upper <- str_to_upper(y)
  return(str_c(y.upper, x, sep = ": "))
}

example.list.4 <- map2(example.list.3, names(example.list.3), upper.name)
str(example.list.3)
str(example.list.4)
```

Exercise 4: Iteration, data.frame manipulation
----------------------------------------------

We know how to add a Year column. We know how to iterate using `map2`. All we
need to do now is put the two things together.

1.  Write a function named `add.year` that takes a data.frame and a vector as
    arguments and returns a data.frame with the vector included as a column.
```{r}

```
2.  Test your function on the first elements of `girls.data` and
    `names(girls.data)` from step 1.
```{r}

```
3.  Use the `map2` function to iterate over `girls.data` and insert the
    corresponding year from `names(girls.data)` using the function you
    wrote in step 1. Do the same for `boys.data`.
```{r}

```

\

\

\

\

\


Exercise 4 prototype 
-------------------------

```{r, results = 'hide', purl = FALSE}
## 1. write function to add year column
add.column <- function(data, value, col.name = "xxxxx") {
  data[[col.name]] <- value
  return(data)
}

## 2. test out the function
str(add.column(girls.data[[1]], value = names(girls.data[1]), col.name = "Year"), max.level = 1)

## 3. map the function to each element of our lists of data.frames
girls.data <- map2(girls.data, names(girls.data), add.column, col.name = "Year")
boys.data <- map2(boys.data, names(boys.data), add.column, col.name = "Year")
```

Final England and Wales data cleanup
====================================

Our next step to is to collapse each list of data.frames into a single
list. We can do that using the `bind_rows` function.

```{r}
library(dplyr)

boys.data <- bind_rows(boys.data)
girls.data <- bind_rows(girls.data)
```

Finally, we want to insert columns indicating location and sex, combine
the boys and girls baby names, and convert the "Year" column to integer.

```{r}
girls.data[["Sex"]] <- "F"
boys.data[["Sex"]] <- "M"
baby.names <- bind_rows(girls.data, boys.data)
baby.names$Year <- as.integer(baby.names$Year)
baby.names$Location <- "England and Wales"
```

Now that our data is cleaned up and formatted, lets take a final look to make
sure everything looks correct.

```{r}
str(baby.names)
```

Get US data (strings, iteration, function recap)
================================================

So far we've read in and cleaned up baby name data from England and Wales. The
next task is to do the same for US baby name data. The US baby name data is
stored in a slightly different way: there are separate files for each Stata, and
each file contains columns indicating the state, sex, year, name, and count.
Note that these files do **not** have names in the first row.

Your mission, should you choose to accept it, is to read in all the US baby name
data and combine them with the England and Wales data we read in earlier. To
succeed you will need to use all your wits and avail yourself of any resources
at your disposal (such as the R help system, <http://stackoverflow.com/>,
discussing with your colleagues, or asking the instructor).

Here is the index page to get you started.

```{r}
baby.names.path <- "example_data/baby_names/US" 
baby.names.url <- str_c(baby.names.path, "index.html", sep = "/")
baby.names.page <- read_lines(baby.names.url)
```

Exercise 5: strings, iteration, functions recap
-----------------------------------------------

You are now a trained R programmer. Put your skills to use as follows

1.  Extract the file names from baby.names.page and combine them with
    `baby.names.path` to create a vector of file locations.
```{r}

```
2.  Iterate over the vector of file names from step one and read each
    file, assigning the result to a list named `us.baby.names`. Note
    that column names are not present in these data files.
```{r}

```
3.  Use `bind_rows` to combine your list of 51 data.frames into a
    single data.frame.
```{r}

```
4.  Use the assignment form of the `names` function to name the columns
    in your US baby names data set "Location", "Sex", "Year",
    and "Count".
```{r}

```
5.  Write a function named `read.linked.files` that wraps steps 1-3.
    Your function should take 3 arguments:

    url
    :   the location of the web page containing data links

    pattern
    :   a regular expression used to extract data links from the page

    reader.fun
    :   a function used to read the files

\

\

\

\

\ 


```{r}
## 1. extract file names
file.names <- na.omit(str_extract(baby.names.page, "[A-Z]+\\.csv"))
file.locations <- str_c(baby.names.path, file.names, sep = "/")

## 2. read in the data
us.baby.names <- map(file.locations,
                     read_csv,
                     col_names = FALSE,
                     na = "",
                     col_types = "ccici")

## 3. collapse list of data.frames into a single data.frame
us.baby.names <- bind_rows(us.baby.names)

## 4. add column names
names(us.baby.names) <- c("Location", "Sex", "Year", "Name", "Count")

## 5. wrap it in a function
read.linked.files <- function(url,
                              pattern = "[A-z0-9]+\\.csv",
                              reader.fun = read_csv) {
  ## extract file names
  file.names <- na.omit(str_extract(read_lines(url), pattern))
  file.locations <- str_c(dirname(url), file.names, sep = "/")
  
  ## read in the data
  the.data <- map(file.locations,
                  reader.fun)
  
  ## collapse list of data.frames into a single data.frame
  return(bind_rows(the.data))
}

## test it out.
test <- read.linked.files("example_data/baby_names/US/index.html",
                          pattern = "[A-Z]+\\.csv",
                          reader.fun = function(x) {
                            read_csv(x, col_names = FALSE, na = "", col_types = "ccici")
                          })

```

Final Data cleanup
------------------

Now that we've downloaded the baby name data from England and Wales and from the
US are ready to do some final data cleanup. First I'm going to drop the "Rank"
variable since it only exists in the England and Wales data, and I don't really
care about it anyway. We can drop elements by assigning `NULL` to them.

```{r}
baby.names[["Rank"]] <- NULL
```

Next we want to combine the two sets of baby names into a single data.frame.
I'll also remove `us.baby.names` since it is quite large and we don't need it
anymore.

```{r}
baby.names <- bind_rows(baby.names, us.baby.names)
rm(us.baby.names)
```

Finally, since the England names are all uppercase and the US names are title
case we need to make them the same.

```{r}
baby.names[["Name"]] <- str_to_lower(baby.names[["Name"]])
```

What can we learn about baby names? (data frame manipulation, graphing)
=======================================================================

Now that we have the data loaded and cleaned up we can use it to answer
interesting questions. For example:

-   How many children were born each year?
-   What are the most popular names overall?
-   How has the popularity of the most popular names changed over time?
-   Has the number of names changed over time?
-   How has the distribution of names changed over time?
-   Has the average length of names changed over time?
-   What are the most popular first letters in baby names? Has this
    changed over time?
-   What were the most popular names in 2015 (the last year for which we
    have data)?
-   How has the popularity of the most popular names in 2015 changed
    over time?

Earlier I mentioned that data.frames are a special kind of list in which each
element must be of the same length. What I perhaps did not emphasize enough is
how important the data.frame concept is in R programming. Because of the
centrality of the data.frame in many R workflows there are many functions
available specifically for manipulating data.frames. In particular, the
*data.table*, *dplyr* and *tidyr* packages provides many useful functions for
working with data.frames. In this section we will use the *dplyr* package answer
some of our questions about baby names.

Number of children born each year
---------------------------------

Now that we have the data read in and we know what is in each column, I want to
know how many children were born each year. To one way to do that is to use the
`group_by` and `summarize` functions from the *dplyr* package.

```{r}
## make up some example data
(example.df <- data.frame(id  = rep(letters[1:4], each = 4),
                          t   = rep(1:4, times = 4),
                          var1 = runif(16),
                          var2 = sample(letters[1:3], 16, replace = TRUE)))
```

Calculate the total sum of the `var1` column.
```{r}
## summarize
summarize(example.df,
          var1 = sum(var1))
```

Calculate the sum of `var1` by `id`.
```{r}
## summarize by id
example.df %>%
  group_by(id) %>%
  summarize(var1 = sum(var1))
```

Calculate the sum of `var1` by `id` and `var2`.
```{r}
## summarize by id and var2
example.df %>%
  group_by(id, var2) %>%
  summarize(var1 = sum(var1))
```

Now that we know how to group by and summarize we can calculate the
number of babies born each year

```{r}
group_by(baby.names, Year) %>%
  summarize(Total = sum(Count))
```

We can also easily generalize this to other groupings, e.g., to
calculate the number of boys and girls born each year:

```{r}
group_by(baby.names, Year, Sex) %>%
  summarize(Total = sum(Count))
```

Most popular names overall
--------------------------

Next I want to find the five most popular girl and boy names overall.

Specifically I want to:

1.  Compute "Percent" column by "Location", "Sex" and "Year".
2.  Average the "Percent" column for each name by "Sex".
3.  Sort names by the averaged Percent.
4.  Select the top 5 in each group.

In order to identify the most popular names it will be useful to learn a
few new tools. You already know how `group_by` and `summarize` work. The
following example shows how `mutate`, `arrange` and `slice` work.

```{r}
## add new columns using mutate
(example.df <- example.df %>%
   mutate(n1 = n(),
          n2 = length(var1),
          var3 = var1/n1))
```

```{r}
## combine with group_by
(example.df <- example.df %>%
   group_by(id) %>%
   mutate(n3 = n(),
          var4 = var1/n3))
```

```{r}
## summarize by id and var2
(example.df.2 <- example.df %>%
   group_by(id, var2) %>%
   summarize(var1 = sum(var1)))
```

```{r}
## sort by var1
example.df.2 %>%
  arrange(var1)
```

```{r}
example.df.2 %>%
  arrange(desc(var1))
```

```{r}
## sort by var1 within id
(example.df.2 <- example.df.2 %>%
   arrange(id, desc(var1)))
```

```{r}
## select the top 2 var1's within each id group
example.df.2 %>%
  group_by(id) %>%
  slice(1:2)
```

Now that we know how to group, mutate, summarize, arrange, and slice we
can use these tools to identify the most popular boy and girl names.

```{r}
## create a percent column
baby.names <- baby.names %>%
  group_by(Location, Year, Sex) %>%
  mutate(Percent = (Count/sum(Count)) * 100)

## collapse across year and location to get the average percent for each each name by sex
baby.names.percent.overall <- baby.names %>%
  group_by(Sex, Name) %>%
  summarize(Percent = mean(Percent))

## arrange the data with the highest average percent first, and select the top 5
(top.names.overall <- baby.names.percent.overall %>%
   arrange(Sex, desc(Percent)) %>%
   group_by(Sex) %>%
   slice(1:5))
```

How has the distribution of names changed over time?
----------------------------------------------------

Finally, I want to examine how the distribution of names has changed
over time. This doesn't require learning any new tools, we just need to
combine the tools we already know to produce the desired result.

```{r}
baby.names.sex.decade <- baby.names %>%
  ## aggregate Year to Decade
  mutate(Decade = str_c(floor(Year/10)*10, "'s")) %>%
  ## calculate Percent by Location, Sex, and Year
  group_by(Location, Sex, Year) %>%
  mutate(Percent = (Count/sum(Count)) * 100) %>%
  ## summarize to the Sex/Decade/Name level
  group_by(Sex, Decade, Name) %>%
  summarize(Count = sum(Count),
            Percent = mean(Percent)) %>%
  ## order the names by overall popularity.
  group_by(Sex) %>%
  mutate(Name = reorder(Name, -1*Percent))

## plot
library(ggplot2)

ggplot(baby.names.sex.decade, aes(x = Name, y = Percent, fill = Sex, color = Sex)) +
  geom_bar(stat = "identity") +
  facet_grid(Decade ~ Sex, scales = "free_x") +
  theme(axis.text.x = element_blank())
```

Exercise 5: Final project
-------------------------

Choose one of these questions

-   ~~How many children were born each year?~~
-   ~~What are the most popular names overall?~~
-   How has the popularity of the most popular names changed over time?
-   Has the number of names changed over time?
-   ~~How has the distribution of names changed over time?~~
-   Has the average length of names changed over time?
-   What are the most popular first letters in baby names? Has this
    changed over time?
-   What were the most popular names in 2015 (the last year for which we
    have data)?
-   How has the popularity of the most popular names in 2015 changed
    over time?

and use the `baby.names` data together with your mastery of R to answer
it.

```{r}

```

Other stuff
===========

Combining similar names
-----------------------

Similar names with different spellings have been intentionally left
alone in these data. If our goals is to examine changes in name
popularity (rather than changing spelling preferences) it might be
useful to combine them. There are several different algorithms for
detecting words that sound alike; many of them (including the *nysiis*
and *metaphone* algorithms) are provided in R by the *phonics* package.
Here is how it works:

```{r}
##install.packages("phonics")
library(phonics)
spelling <- c(
"colour",   "color",       
"flavour",   "flavor",   
"behaviour", "behavior", 
"harbour",   "harbor",   
"honour",    "honor",    
"humour",    "humor",    
"labour",    "labor",    
"neighbour", "neighbor", 
"rumour",    "rumor")

(spelling.data <- data.frame(original = spelling, Sounds.like = nysiis(spelling, maxCodeLen = 15)))
```

Once we have identified which words are the same (but spelled
differently) we can combine those words.

```{r}
(spelling.data <- spelling.data %>%
   group_by(Sounds.like) %>%
   mutate(Alternative.spellings = str_c(sort(original), collapse = "/")))
```

We can use this technique to identify and group names with alternative
spellings. Since any one algorithm tends to group two aggressively we'll
require multiple algorithms to agree.

```{r}
library(phonics)
baby.names <- baby.names %>%
  group_by(Sex) %>%
  mutate(Sounds.like = str_c(nysiis(Name, maxCodeLen = 10),
                             metaphone(Name, maxCodeLen = 10),
                             sep = "/")) %>%
  group_by(Sex, Sounds.like) %>%
  mutate(Alternative.spellings = str_c(sort(unique(Name)), collapse = "/")) %>%
  ungroup()
str(baby.names)
```

"starts with" and "ends with"
-----------------------------

Not only do the most popular names change over time, but the most
popular first and last letters also display temporal trends. We can
extract the first/last n letters using `str_sub`, like this:

```{r}
month.name
str_sub(month.name, 1, 1)
str_sub(month.name, 1, 3)
str_sub(month.name, -1, -1)
str_sub(month.name, -3, -1)
```

We can use this to create new columns with just the first/last **n**
letters of each name

```{r}

baby.names[["First.1"]] <- str_sub(baby.names[["Name"]], 1, 1)
baby.names[["First.2"]] <- str_sub(baby.names[["Name"]], 1, 2)
baby.names[["First.3"]] <- str_sub(baby.names[["Name"]], 1, 3)
baby.names[["Last.1"]] <- str_sub(baby.names[["Name"]], -1, -1)
baby.names[["Last.2"]] <- str_sub(baby.names[["Name"]], -2, -1)
baby.names[["Last.3"]] <- str_sub(baby.names[["Name"]], -3, -1)
```

Length of names
---------------

The `str_length` function can be used to count the number of letters in
each name

```{r}
baby.names$Name.length <- str_length(baby.names$Name)
```

What else?
==========

If there is anything else you want to learn how to do, now is the time
to ask!

Go forth and code!
==================

You now know everything you could possibly want to know about R. OK
maybe not! But you do know how to manipulate character strings with
regular expressions, write your own functions, iterate, inspect and
modify attributes, and extract and replace object elements. There's a
lot more to learn, but that's a pretty good start. As you go forth and
write your own R code here are some resources that may be helpful.

Additional reading and resources
================================

-   Learn from the best: <http://adv-r.had.co.nz/>;
    <http://r4ds.had.co.nz/>
-   R documentation: <http://cran.r-project.org/manuals.html>
-   Collection of R tutorials:
    <http://cran.r-project.org/other-docs.html>

-   R for Programmers (by Norman Matloff, UC--Davis)

<http://heather.cs.ucdavis.edu/~matloff/R/RProg.pdf>

-   Calling C and Fortran from R (by Charles Geyer, UMinn)

<http://www.stat.umn.edu/~charlie/rc/>

-   State of the Art in Parallel Computing with R (Schmidberger et al.)

<http://www.jstatso>|.org/v31/i01/paper

-   Institute for Quantitative Social Science: <http://iq.harvard.edu>
-   Research technology consulting: <http://projects.iq.harvard.edu/rtc>

Epilogue: Things that may surprise you
======================================

There are an unfortunately large number of surprises in R programming.
Some of these "gotcha's" are common problems in other languages, many
are unique to R. We will only cover a few -- for a more comprehensive
discussion please see
<http://www.burns-stat.com/pages/Tutor/R_inferno.pdf>

### Floating point comparison

Floating point arithmetic is not exact:

```{r}
.1 == .3/3
```

Solution: `use all.equal()`:

```{r}
all.equal(.1, .3/3)
```

### Missing values

R does not exclude missing values by default -- a single missing value
in a vector means that many thing are unknown:

```{r}
x <- c(1:10, NA, 12:20)
c(mean(x), sd(x), median(x), min(x), sd(x))
```

NA is not equal to anything, not even NA

```{#NAcomp .r}
NA == NA
```

Solutions: use `na.rm = TRUE` option when calculating, and is.na to test
for missing

### Automatic type conversion

Automatic type conversion happens a lot which is often useful, but makes
it easy to miss mistakes

```{r}
# combining values coereces them to the most general type
(x <- c(TRUE, FALSE, 1, 2, "a", "b"))
str(x)

# comparisons convert arguments to most general type
1 > "a"
```

Maybe this is what you expect... I would like to at least get a warning!

### Optional argument inconsistencies

Functions you might expect to work similarly don't always:

```{r}
mean(1, 2, 3, 4, 5)*5
sum(1, 2, 3, 4, 5)
```

Why are these different?!?

```{r}
args(mean)
args(sum)
```

Ouch. That is not nice at all!

### Trouble with Factors

Factors sometimes behave as numbers, and sometimes as characters, which
can be confusing!

```{r}
(x <- factor(c(5, 5, 6, 6), levels = c(6, 5)))

str(x)

as.character(x)
# here is where people sometimes get lost...
as.numeric(x)
# you probably want
as.numeric(as.character(x))
```

Feedback
--------

-   Help Us Make This Workshop Better!
-   Please take a moment to fill out a very short feedback form
-   These workshops exist for you -- tell us what you need!
-   <http://tinyurl.com/RprogrammingFeedback>

[^1]: In general you [should not use string manipulation tools to parse
    html](http://stackoverflow.com/questions/1732348/regex-match-open-tags-except-xhtml-self-contained-tags/1732454#1732454),
    but in this case the html is simple enough that a regular expression
    based approach is fine.
